---
title: "competition1"
output: html_document
date: "2025-05-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##step1 import data

```{r cars}
toenailData <- read.table("toenail.txt",header=T)
```

## discussing variables (hierachical/correlat?)
longitudinal

## step3 decide which variables are factors 

```{r pressure, echo=FALSE}
toenailData$id<-as.factor(toenailData$id)
toenailData$treat<-as.factor(toenailData$treat)
toenailData$center<-as.factor(toenailData$center)
toenailData$time<-as.factor(toenailData$time)
levels(toenailData$treat)<-c("1","2")
```

## step4 plot nailbig vs. id and check on object variability 

```{r}
allNail<-unique(toenailData$id)
nNail<-length(allNail)

plot(as.numeric(toenailData$id),toenailData$nailbig,type="n",xlab="ID",
     ylab="Nailbig",main="Nailbig by ID", ylim =c(-0,25))


for (i in 1:nNail){
  tmp<-toenailData[toenailData$id==allNail[i],]
  tmp$id<-as.numeric(tmp$id)
  points(allNail[i],mean(tmp$nailbig),pch=1)
}  


```
## step 5 plot nailbig vs. time
```{r}
allNail<-unique(toenailData$id)
nNail<-length(allNail)

toenailData$time <- as.numeric(toenailData$time)

plot(toenailData$nailbig~toenailData$time,type="n",xlab="time",
     ylab="Nailbig",main="Nailbig by time")


for (i in 1:nNail){
  tmp<-toenailData[toenailData$id==allNail[i],]
  lines(tmp$nailbig ~ tmp$time,pch=1)
  points(tmp$nailbig ~ tmp$time,pch=1)
} 
```

## step 6 comparing two treatments - independent t-test 

```{r}

t.test(toenailData$nailbig ~ toenailData$treat)
```


##step 7 visualize effect in time of the treatment 1 and treatment 2
```{r}

allNail<-unique(toenailData$id)
nNail<-length(allNail)

toenailData$time <- as.numeric(toenailData$time)

plot(toenailData$nailbig~toenailData$time,type="n",xlab="time",
     ylab="Nailbig",main="Nailbig by time")


for (i in 1:nNail){
  tmp<-toenailData[toenailData$id==allNail[i],]
  points(tmp$nailbig ~ tmp$time, col = tmp$treat,pch=1)
  lines(tmp$nailbig ~ tmp$time, col = tmp$treat,pch=1)
} 
```

## step 8 
starting with linear model bacause nailbig grew with time
ANOVA identifies which factors are important - independent variables - time and treatment 
interaction model: whether time and treatment interacts to influence nailbig values 
selecting the fixed and random effects to see modeling 

