---
title: "Competition 2 Multilevel data"
author: "Roma Siugzdaite"
date: "2025-05-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Competition II.

**Data **
From SPSS dataset Galo.sav

school - school id
sex - (1 girl, 2 boy)
galo - atchievement score on Galo test (Inteligence test)
advice - advice of the teacher (0 - no subsequent school, 1- lowest, 5 - highest)
feduc - fathers education (1 - lowest, 9- highest, 4 - missing)
meduc - mathers education (1 - lowest, 9- highest, 3 - missing)
facc - father occupation status (1 - lowest, 6 - highest)
denom - denomination school ( 1- protestant, 2 - non-denominational, 3 -catholic)
nmis - nr of missing values (0 - no, 1 - 4)



**Research questions:**

0. Import the data and check the variables.
1. Is any difference in Galo score between genders? Visuals and stats.
2. Does Galo score depend on father education? Visuals and stats.
3. Does Galo score depend on mother education? Visuals and stats.
4. How the advice of teachers depend on the Galo scores? Visuals and stats.
5. Does the Galo scores depend on the denomination of the school? Visuals and stats.
6. Are gender and type of school are important for Galo score? Visuals and model.
7. What random effect(s) could be important for the model in question 6?
8. How big is the variance explained and ICC? 
9. Interpret your results!


```{r, ECHO = TRUE}

library(haven)
Galo <- read_sav("/Users/rsiugzdaite/Documents/JOB/UANTWERPEN/FLAMES/Multilevel/Data/Competition/Galo.sav")
head(Galo)

library(ggplot2)
library(ggpubr)
library(sjPlot)
library(lme4)

```

## Including Plots

```{r , echo=FALSE}

Galo$sex <- as.factor(Galo$sex)
Galo$advice <- as.factor(Galo$advice)
Galo$feduc <- as.factor(Galo$feduc)
Galo$meduc <- as.factor(Galo$meduc)
Galo$focc <- as.factor(Galo$focc)
Galo$denom <- as.factor(Galo$denom)
Galo$nmis <- as.factor(Galo$denom)

plot(galo ~ sex, data = Galo, col = sex)

plot(galo ~ school, data = Galo, col = sex)

plot(galo ~ advice, data = Galo, col = advice)

plot(galo ~ feduc, data = Galo)

plot(galo ~ meduc, data = Galo)

plot(galo ~ focc, data = Galo)

plot(galo ~ denom, data = Galo)

plot(galo ~ nmis, data = Galo)

Galo$school <- as.factor(Galo$school)

plot(galo ~ school, data = Galo)

```

1. Is any difference in Galo score beween genders?

```{r}

t.test(galo ~ sex, data = Galo)

result <- lm(galo ~ sex, data = Galo)
summary(result)
anova(result)

bxp <- ggboxplot(Galo, x = "sex", y = "galo",
                 color = "sex", palette = "jco")
bxp


```
Answer: 


2. Does Galo score depend on fathers education? 

```{r}

result <- lm(galo ~ feduc, data = Galo)
summary(result)
anova(result)


bxp <- ggboxplot(Galo, x = "feduc", y = "galo",
                 color = "feduc", palette = "jco")
bxp

bxp <- ggboxplot(Galo, x = "feduc", y = "galo",
                 color = "sex", palette = "jco")
bxp

```

Answer: 



3. Does Galo score depend on mothers education? 


```{r}

result <- lm(galo ~ meduc, data = Galo)
summary(result)
anova(result)


bxp <- ggboxplot(Galo, x = "meduc", y = "galo",
                 color = "meduc", palette = "jco")
bxp

bxp <- ggboxplot(Galo, x = "meduc", y = "galo",
                 color = "sex", palette = "jco")
bxp

```

Answer:


4. How the advice of teachers depend on the Galo scores?


```{r}

result <- lm(galo ~ advice, data = Galo)
summary(result)
anova(result)

bxp <- ggboxplot(Galo, x = "advice", y = "galo",
                 color = "advice", palette = "jco")
bxp

bxp <- ggboxplot(Galo, x = "advice", y = "galo",
                 color = "sex", palette = "jco")
bxp
```

Answer:


5. Does the Galo scores depend on the denomination of the school?

```{r}

result <- lm(galo ~ denom, data = Galo)
summary(result)
anova(result)

bxp <- ggboxplot(Galo, x = "denom", y = "galo",
                 color = "denom", palette = "jco")
bxp

bxp <- ggboxplot(Galo, x = "denom", y = "galo",
                 color = "sex", palette = "jco")
bxp
```

6. 

```{r}

result <- lm(galo ~ sex * denom , data = Galo)
summary(result)
anova(result)

tab <- tab_model(result)
tab


bxp <- ggboxplot(Galo, x = "sex", y = "galo",
                 color = "denom", palette = "jco")
bxp




```

Answer:


```{r}
result <- lmer(galo ~ sex*denom + (1 |school) + (1|advice), data = Galo)
summary(result)
anova(result)

tab <- tab_model(result)
tab


result <- lmer(galo ~ sex + denom +  (1 |school) + (1|advice), data = Galo)
summary(result)
anova(result)

tab <- tab_model(result)
tab

```


Answer:

