---
title: "Dictionary for MLM"
author: "RSiugzdaite"
date: "2024-09-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Models in R

Let's download a dataset that we can practice some models

```{r }
library(lme4)
library(Matrix)

setwd("~/Documents/JOB/UANTWERPEN/FLAMES/Multilevel/Data")
siiData <- read.table("SII.txt",header=T)
siiData$schoolid<-as.factor(siiData$schoolid)
siiData$classid<-as.factor(siiData$classid)
siiData$childid<-as.factor(siiData$childid)
siiData$sex<-as.factor(siiData$sex)
levels(siiData$sex)<-c("male","female")
```

## t-test

To use t-test we need independent measurement.For example between male and female.

```{r, echo=FALSE}
plot(mathgain ~ sex, data = siiData)
t.test(siiData$mathgain ~ siiData$sex)

```
We see there is no differences between girls and boys in mathgain.

There are no data of measurements before and after, so there is no need to use paired t-test here. 

But in case if we would be using paired t-test, the equation would be the same: t.test(Y ~ X,  paired = TRUE)

In case of ses

```{r}

siiData$yearstea <- as.factor(siiData$yearstea)
plot(mathgain ~ yearstea, data = siiData)
fit <- lm(mathgain ~ yearstea, data = siiData)
anova(fit)
summary(fit)

```
Linear regression is about two continues variables, their linear relationship.

```{r}

plot(mathgain ~ ses, data = siiData)
fit <- lm(mathgain ~ ses, data = siiData)
summary(fit)
plot(fit)

```


```{r}

allSchool<-unique(siiData$schoolid)
nSchool<-length(allSchool)

plot(as.numeric(siiData$schoolid),siiData$mathgain,type="n",xlab="School ID",ylab="Mathgain",main="Mathgain by school")

for (i in 1:nSchool){
  tmp<-siiData[siiData$schoolid==allSchool[i],]
  tmp$schoolid<-as.numeric(tmp$schoolid)
  
  #mean for the males	
  tmpMale<-tmp[tmp$sex=="male",]
  points(allSchool[i],mean(tmpMale$mathgain),pch=16)
  
  #mean for the females
  tmpFemale<-tmp[tmp$sex=="female",]
  points(allSchool[i],mean(tmpFemale$mathgain),pch=1)
}

model<-lme(mathgain~sex,random=~1|schoolid,data=siiData)
anova.lme(model)
summary(model)

#the same thing in a different way



```

